\documentclass[msc]{on} % para dissertação de mestrado
%\documentclass[dsc]{on} % para Tese de doutorado
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
\usepackage{color}
\usepackage{verbatim}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{multirow}
\usepackage[utf8]{inputenc}
\usepackage{xcolor}
\usepackage{soul}
\setstcolor{red} % Sets the strikethrough color to red
% para mostrar os links coloridos
\usepackage{hyperref}
\hypersetup{%unicode=true,bookmarks=false,backref=false,
colorlinks,linkcolor=blue,citecolor=red,urlcolor=magenta,menucolor=cyan}   

%\usepackage{indentfirst}
%\usepackage{parskip}

% modifca as margens padrao
%\usepackage{geometry}  
%\geometry{top=2.5cm,bottom=2.5cm,inner=2.5cm,outer=1.5cm}  

% figuras multiples com (a), (b), (c), etc
%\usepackage{subfigure} 

% inclui as subsubsubsecoes no sumario
%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}   

% listas de simbolos e abreviacoes
% para criar as listas de abreviaturas e/ou símbolos, chamar ao longo do texto as macros 
% SIGLA\abbrev{SIGLA}{definicão de SIGLA} ou 
% SIMBOLO\symbl{SIMBOLO}{definicão de SIMBOLO}
% compilar as listas com os comandos
%     makeindex -s on.ist -o filename.lab filename.abx
%     makeindex -s on.ist -o filename.los filename.syx
%\makelosymbols
%\makeloabbreviations

%%%%%%%%%%%%
\begin{document}

% titulo em portugues e ingles
\title{Pipeline para Detecção Automatizada de Ocultações Estelares em Curvas de Luz com Técnicas de Machine Learning}
\foreigntitle{Automated Pipeline for Detecting Stellar Occultations in Light Curves Using Machine Learning}

% autor e orientador e coorientador(es)
% exemplo: Jose Maria da Silva => Nome: Jose Maria da - Sobrenome: Silva
%          Joao Santos Junior => Nome Joao - Sobrenome: Santos Junior
% usar abreviatura das instituicoes
% exemplo: OV/UFRJ
%          UDELAR, Uruguai
\author{Thiago}{Laidler Vidal Cunha}
\advisor{Dr.}{J\'ulio}{Camargo}{Observat\'orio Nacional}
%\coadvisor{Prof.}{Nome}{Sobrenome}{Instituicao}
%\coadvisor{Prof.}{Nome}{Sobrenome}{Instituicao}

% o codigo cutter deve corresponder às primeiras letras do sobrenome do autor
% consultar a tabela em https://www.icmc.usp.br/institucional/estrutura-administrativa/biblioteca/servicos/cutter
% exemplo: sobrenome Santos, letra S, cutter 237
%          sobrenome Costa, letra C, cutter 837
\cutter{837}

% o codigo CDU sera fornecido pela biblioteca posteriormente
\cducode{000.000.000}

% membros da banca - nao incluir orientador nem coorientador(es)
\examiner{Dr.}{Nome Sobrenome}{Instituicao}
\examiner{Profa.}{Nome Sobrenome}{Instituicao}
\examiner{Dra.}{Nome Sobrenome}{Instituicao}
\examiner{Prof.}{Nome Sobrenome}{Instituicao}
%\examiner{Prof.}{Nome Sobrenome}{Instituição}

% COPAA ou COGEO
\department{COPAA}

% data da defesa - o dia podera ser informado depois
\date{10}{06}{2026}

% inserir ate 6 palavras chave
\keyword{Ocultação Estelar}
\keyword{Fotometria}
\keyword{Aprendizado de Máquina}
\keyword{Machine Learning}
\keyword{Big Data}
\keyword{Ciencia de Dados}

\maketitle

%%%%%%%%%%%%%%%%%%%
%\frontmatter
%\dedication{%
%Dedicat\'oria (opcional).
%}

%%%%
\acknowledgments
\input{agradecimentos}

%%%%
\cleardoublepage
\thispagestyle{plain}
%\epigraph{Ep\'igrafe (opcional).}
%\vspace*{\fill}


%------------------------------
%   EPÍGRAFE (ESTA PÁGINA É OPCIONAL)
%------------------------------


\pagestyle{empty} % No headers or footers for the following pages

\null\vfill\vfill\vfill\vfill\vfill\vfill\vfill % Add some space to move the quote down the page a bit

\begin{flushright}
{\normalsize 

\textit{"O primeiro princípio é que você não deve enganar a si mesmo - e você é a pessoa mais fácil de se enganar."}

- Richard Feynman

}
\end{flushright}

\vfill\vfill\vfill\null % Add some space at the bottom to position the quote just right

\clearpage

%%%%
\begin{abstract}
Entre as técnicas disponíveis para a caracterização física de asteroides, Centauros e objetos transnetunianos, as ocultações estelares destacam-se pela elevada precisão. Nesses eventos, um corpo do Sistema Solar bloqueia temporariamente a luz de uma estrela distante para um dado observador. A análise da curva de luz resultante permite inferir tamanho, forma projetada e, em casos favoráveis, atmosfera e anéis do corpo ocultante, com resolução espacial da ordem de quilômetros. O crescimento do volume de dados produzido por redes de telescópios e catálogos como o Gaia, no entanto, tornou a triagem manual de todas as curvas um problema relevante e motivou o desenvolvimento de métodos automatizados de detecção. Esta dissertação apresenta uma \textit{pipeline} completa para detecção automatizada de ocultações estelares em curvas de luz, baseada em técnicas de aprendizado de máquina. O problema é formulado como classificação binária: cada curva é descrita por um vetor de características numéricas (\textit{features}) e classificada como positiva (ocultação presente) ou negativa (ausência de evento). Quatro classificadores foram treinados e comparados: Regressão Logística, Random Forest, XGBoost e CatBoost. O conjunto de dados reúne 1693 curvas (802 positivas e 891 negativas). Seis experimentos avaliaram o impacto da seleção de \textit{features} e a generalização para dados exclusivamente reais. Os resultados demonstram desempenho elevado e consistente: F1-score entre 0,978 e 0,994 e AUC-ROC superior a 0,998 no conjunto de teste. A \textit{pipeline} permite ainda o ajuste do limiar de decisão em pós-processamento, operando em regime de alta sensibilidade sem necessidade de retreinamento. Nesse contexto, o custo de perder um evento genuíno (falso negativo) supera o de revisar um alarme falso (falso positivo). Como validação em dados reais, utilizou-se a ocultação pelo objeto transnetuniano (50000) Quaoar, não incluída no conjunto de treinamento. A aplicação da ferramenta permitiu avaliar a identificação de ocultações em uma curva de luz complexa. Embora os sinais produzidos pelos anéis finos apresentem probabilidades absolutas baixas, estas são cerca de duas ordens de grandeza superiores às atribuídas a trechos de ruído semelhantes. Como consequência, um ajuste simples do limiar é suficiente para recuperar esses eventos sutis sem introduzir falsos alarmes. Os resultados demonstram o potencial da ferramenta para revisitar curvas de luz complexas em busca de eventos mais sutis ainda não identificados numa primeira análise. Assim, este trabalho contribui com uma ferramenta reproduzível para a triagem e priorização de curvas de luz em estudos de ocultações estelares.
\end{abstract}

%%%%
\begin{foreignabstract}
Among the techniques available for the physical characterization of asteroids, Centaurs and trans-Neptunian objects, stellar occultations stand out for their high precision. In such events, a Solar System body temporarily blocks the light of a distant star as seen by a given observer. Analysis of the resulting light curve allows one to infer the size, projected shape and, in favorable cases, the atmosphere and rings of the occulting body, with spatial resolution on the order of kilometers. The growing data volume produced by telescope networks and catalogs such as Gaia has, however, made the manual screening of all light curves a relevant problem and motivated the development of automated detection methods. This dissertation presents a complete pipeline for the automated detection of stellar occultations in light curves, based on machine learning techniques. The problem is cast as binary classification: each curve is described by a vector of numerical features and classified as positive (occultation present) or negative (no event). Four classifiers were trained and compared: Logistic Regression, Random Forest, XGBoost and CatBoost. The dataset comprises 1693 light curves (802 positive and 891 negative). Six experiments assessed the impact of feature selection and the generalization to exclusively real data. The results demonstrate high and consistent performance: F1-scores between 0.978 and 0.994 and AUC-ROC above 0.998 on the test set. The pipeline also supports decision-threshold tuning as a post-processing step, operating in a high-recall regime without the need for retraining. In this context, the cost of missing a genuine event (false negative) outweighs that of reviewing a false alarm (false positive). As a validation on real data, the occultation by the trans-Neptunian object (50000) Quaoar, not included in the training set, was used. Applying the tool made it possible to assess the identification of occultations in a complex light curve. Although the signals produced by the thin rings have low absolute probabilities, these are about two orders of magnitude higher than those assigned to similar noise segments. As a consequence, a simple threshold adjustment is enough to recover these subtle events without introducing false alarms. The results demonstrate the potential of the tool to revisit complex light curves in search of subtler events that a first analysis leaves unidentified. This work thus contributes a reproducible tool for the screening and prioritization of light curves in stellar-occultation studies.
\end{foreignabstract}

%%%%
\listoffigures
\listoftables
%\printlosymbols
%\printloabbreviations
\tableofcontents

%%%%%%%%%%%%%%%%%%%
\mainmatter

\chapter{Introdu\c c\~ao}

\input{introducao}

\chapter{Ocultações Estelares e Processamento de Dados}

\input{capitulo2}

\chapter{Modelos e algoritmos para Machine Learning}

\input{capitulo3}

\chapter{Metodologia: Pipeline de Detecção}

\input{capitulo4}

\chapter{Resultados}

\input{capitulo5}

\chapter{Conclusões}
\label{cap:conclusoes}

\input{capitulo6}

%%%%%%%%%%%%%%%%%%%
\backmatter

% lista de referencia BibTeX: é necessario criar o arquivo filename.bib
% as referencias sao invocadas no texto com os macros \citet{key} ou \citep{key}. etc.
% criar a lista de referencias com o comando
%    bibtex filename.aux
%\bibliographystyle{on}
%\bibliography{filename}

% lista de referencias manual
\begin{thebibliography}{100}

% - Mais citados e fundamentos de ML/estatística -
\bibitem[Geron (2019)]{geron2019maos} G\'ERON, Aur\'elin. \textit{M\~aos \`a Obra: Aprendizado de M\'aquina com Scikit-Learn e TensorFlow}. 2. ed. Rio de Janeiro: Alta Books, 2019.

\bibitem[Levada (2026)]{Levada2026} LEVADA, Alexandre L. M. \textit{Manual completo para aprendizado de m\'aquina: algoritmos para classifica\c{c}\~ao supervisionada, agrupamento de dados e redu\c{c}\~ao de dimensionalidade}. Departamento de Computa\c{c}\~ao, Universidade Federal de S\~ao Carlos (UFSCar), 16 jan. 2026.

\bibitem[Cazeneuve et al. (2023)]{Cazeneuve2023} CAZENEUVE, Dorian; MARCHIS, Franck; BLACLARD, Guillaume; DALBA, Paul A.; MARTIN, Victor; ASENCIO, Joe. ``ODNet: A Convolutional Neural Network for Asteroid Occultation Detection''. \emph{The Astronomical Journal}, 165(11), 2023.

\bibitem[Wall \& Jenkins (2003)]{Wall2003} WALL, J. V.; JENKINS, C. R. \textit{Practical Statistics for Astronomers}. Cambridge University Press, 2003.

\bibitem[Morettin \& Bussab (2010)]{Morettin2010} MORETTIN, Pedro A.; BUSSAB, Wilton de O. \textit{Estatística Básica}. 6. ed. Revista e atualizada. São Paulo: Saraiva, 2010.

\bibitem[Deisenroth et al. (2019)]{Deisenroth2019} DEISENROTH, Marc Peter; FAISAL, A. Aldo; ONG, Cheng Soon. \textit{Mathematics For Machine Learning}. Cambridge University Press, 2019. (Draft, October 15, 2019).

\bibitem[Knieling et al. (2024)]{Knieling2024} KNIELING, Bastian; SCHINDLER, Karsten; SICKAFOOSE, Amanda A.; PERSON, Michael J.; LEVINE, Stephen E.; KRABBE, Alfred. ``Stellar Occultations in the Era of Data Mining and Modern Regression Models: Using Gaussian Processes to Analyze Light Curves and Improve Predictions''. \emph{The Planetary Science Journal}, 5(104), 2024. \url{https://doi.org/10.3847/PSJ/ad3819} [1]

\bibitem[Fraser et al. (2024)]{Fraser2024} FRASER, Wesley C.; PORTER, Simon B.; PELTIER, Lowell; KAVELAARS, J. J.; VERBISCER, Anne J.; et al. ``Candidate Distant Trans-Neptunian Objects Detected by the New Horizons Subaru TNO Survey''. \emph{arXiv e-prints}, arXiv:2407.21142v1, 2024.

% - Ocultações, curvas de luz e redução de dados -
\bibitem[Gomes-Júnior et al. (2022)]{SORA2022} GOMES-JÚNIOR, A. R.; MORGADO, B. E.; BENEDETTI-ROSSI, G.; BOUFLEUR, R. C.; ROMMEL, F. L.; BANDA-HUARCA, M. V.; KILIC, Y.; BRAGA-RIBAS, F.; SICARDY, B. ``SORA: Stellar occultation reduction and analysis''. \emph{Monthly Notices of the Royal Astronomical Society}, 511(1), 1167--1181, 2022. \url{https://doi.org/10.1093/mnras/stac032}.

\bibitem[Roques et al. (1987)]{Roques1987} ROQUES, F.; MONCUQUET, M.; SICARDY, B. ``Stellar occultations by small bodies - Diffraction effects''. \emph{The Astronomical Journal}, 93, 1549--1558, 1987. \url{https://doi.org/10.1086/114438}.

\bibitem[Sicardy et al. (2011)]{Sicardy2011} SICARDY, B.; ORTIZ, J. L.; ASSAFIN, M.; et al. ``A Pluto-like radius and a high albedo for the dwarf planet Eris from an occultation''. \emph{Nature}, 478(7370), 493--496, 2011. \url{https://doi.org/10.1038/nature10550}.

\bibitem[Sicardy et al. (2016)]{Sicardy2016} SICARDY, B.; TALBOT, J.; MEZA, E.; CAMARGO, J. I. B.; et al. ``Pluto's atmosphere from the 2015 June 29 ground-based stellar occultation at the time of the New Horizons flyby''. \emph{The Astrophysical Journal Letters}, 819(2), L38, 2016. \url{https://doi.org/10.3847/2041-8205/819/2/L38}.

\bibitem[Assafin (2023)]{Assafin2023} ASSAFIN, M. \textit{Differential aperture photometry and digital coronagraphy with PRAIA}. Planetary and Space Science, v. 239, p. 105816, dez. 2023. DOI: 10.1016/j.pss.2023.105816.

\bibitem[Braga-Ribas et al. (2014)]{BragaRibas2014} BRAGA-RIBAS, F.; SICARDY, B.; ORTIZ, J. L.; SNODGRASS, C.; ROQUES, F.; et al. ``A ring system detected around the Centaur (10199) Chariklo''. \emph{Nature}, 508(7494), 72--75, 2014. \url{https://doi.org/10.1038/nature13155}.

\bibitem[Assafin et al. (2011)]{AssafinPRAIA} ASSAFIN, M.; MARTINS, R.; CAMARGO, J.; ANDREI, A.; NETO, D.; BRAGA-RIBAS, F. ``PRAIA - Platform for Reduction of Astronomical Images Automatically''. 85--88, 2011.

\bibitem[Assafin et al. (2023)]{Umbriel2020} ASSAFIN, M.; SANTOS-FILHO, S.; MORGADO, B. E.; GOMES-JÚNIOR, A. R.; SICARDY, B.; MARGOTI, G.; BENEDETTI-ROSSI, G.; BRAGA-RIBAS, F.; LAIDLER, T.; CAMARGO, J. I. B.; VIEIRA-MARTINS, R.; et al. ``Kilometer-precise (UII) Umbriel physical properties from the multichord stellar occultation on 2020 September 21''. \emph{Monthly Notices of the Royal Astronomical Society}, 526(4), 6193–6204, 2023. \url{https://doi.org/10.1093/mnras/stad3093}.

\bibitem[Ortiz et al. (2023)]{Chiron2023} ORTIZ, J.; PEREIRA, Chrystian; SICARDY, Bruno; BRAGA-RIBAS, Felipe; TAKEY, Ali; FOUAD, A.; SHAKER, A.; KASPI, S. ``Changing material around (2060) Chiron revealed by an occultation on December 15, 2022''. \emph{Astronomy \& Astrophysics}, 676, 2023. \url{https://doi.org/10.1051/0004-6361/202347025}.

\bibitem[Pereira et al. (2023)]{Quaoar2023} PEREIRA, C. L.; SICARDY, B.; MORGADO, B. E.; BRAGA-RIBAS, F.; FERNÁNDEZ-VALENZUELA, E.; SOUAMI, D.; HOLLER, B. J.; BOUFLEUR, R. C.; MARGOTI, G.; ASSAFIN, M.; ORTIZ, J. L.; et al. ``The two rings of (50000) Quaoar''. \emph{Astronomy \& Astrophysics}, 673, L4, 2023. \url{https://doi.org/10.1051/0004-6361/202346365}.

\bibitem[Morgado et al. (2023)]{Morgado2023} MORGADO, B. E.; SICARDY, B.; BRAGA-RIBAS, F.; ORTIZ, J. L.; SALO, H.; VACHIER, F.; DESMARS, J.; PEREIRA, C. L.; SANTOS-SANZ, P.; SFAIR, R.; DE SANTANA, T.; et al. ``A dense ring of the trans-Neptunian object Quaoar outside its Roche limit''. \emph{Nature}, 614(7947), 239--243, 2023. \url{https://doi.org/10.1038/s41586-022-05629-6}.

\bibitem[Gomes-Ferrante \& Braga-Ribas (2023)]{Simulator} GOMES-FERRANTE, Wellington; BRAGA-RIBAS, Felipe. ``Stellar Occultation Simulator: application to Planet 9''. \emph{Eur. Phys. J. Spec. Top.}, 232(18-19), 3113--3118, 2023. \url{https://www.springernature.com/gp/researchers/text-and-data-mining}.

\bibitem[Fornasier et al. (2013)]{Fornasier2013} FORNASIER, Sonia; LELLOUCH, Emmanuel; MÜLLER, Thomas G.; SANTOS-SANZ, Pablo; PANUZZO, Pasquale; KISS, Csaba; LIM, Taeho; MOMMERT, Michael; BOCKELÉE-MORVAN, Dominique; VILENIUS, Essi; STANSBERRY, John; TOZZI, Gian Paolo; MOTTOLA, Stefano; DELSANTI, Audrey; CROVISIER, Jacques; DUFFARD, René; HENRY, Frédéric; LACERDA, Pedro; BARUCCI, Antonella; GICQUEL, Arnaud. \textit{TNOs are Cool: A survey of the trans-Neptunian region VIII. Combined Herschel PACS and SPIRE observations of nine bright targets at 70--500 $\mu$m}. Astronomy \& Astrophysics, v. 555, p. A15, jul. 2013. DOI: 10.1051/0004-6361/201321329.

% - Contexto, ML em ciência e métricas -
\bibitem[Gomes (2009)]{Tsiganis2009} GOMES, Rodney da Silva. \textit{On the origin of the Kuiper belt}, 2009.

\bibitem[Zhu et al. (2022)]{Zhu2022} ZHU, Qinglin; SUN, Mingchen; DONG, Xiang; ZHU, Pengfei. ``Design and Simulation of Stellar Occultation Infrared Band Constellation''. \emph{Remote Sens.}, 14(3327), 2022. \url{https://doi.org/10.3390/rs14143327}

\bibitem[Morbidelli et al. (2024)]{Morbidelli2024} MORBIDELLI, Alessandro; MARROCCHI, Yves; ALI AHMAD, Adnan; BHANDARE, Asmita; CHARNOZ, S\'ebastien; COMMER\c{C}ON, Beno\^it; DULLEMOND, Cornelis P.; GUILLOT, Tristan; HENNEBELLE, Patrick; LEE, Yueh-Ning; LOVASCIO, Francesco; MARSCHALL, Raphael; MARTY, Bernard; MAURY, Ana\"elle; TAMAMI, Okamoto. \textit{Formation and evolution of a protoplanetary disk: Combining observations, simulations, and cosmochemical constraints}. Astronomy \& Astrophysics, v. 691, p. A147, nov. 2024. DOI: 10.1051/0004-6361/202451388.

\bibitem[López (2020)]{lopez2020ml} LÓPEZ DE PRADO, Marcos M. \textit{Machine Learning for Asset Managers}. Cambridge: Cambridge University Press, 2020. (Elements in Quantitative Finance).

\bibitem[Carleo et al. (2019)]{CarleoMLPhysSci} CARLEO, Giuseppe; CIRAC, Ignacio; CRANMER, Kyle; DAUDET, Laurent; SCHULD, Maria; TISHBY, Naftali; VOGT-MARANTO, Leslie; ZDEBOROVÁ, Lenka. \textit{Machine learning and the physical sciences}, 2019. (Review paper excerpts)

\bibitem[Sudjianto \& Liu (2025)]{AUC-curve} SUDJIANTO, Agus; LIU, Alice J. ``Decomposing Global AUC into Cluster-Level Contributions for Localized Model Diagnostics''. \emph{arXiv preprint}, arXiv:2508.07495, 2025. \url{https://arxiv.org/abs/2508.07495}.

\bibitem[Gryak et al. (2018)]{gryak2018conjugacy} GRYAK, J.; HARALICK, R. M.; KAHROBAEI, D. \textit{Solving the Conjugacy Decision Problem via Machine Learning}. Experimental Mathematics, v. 29, n. 1, p. 66–78, 2018. Disponível em: \url{https://doi.org/10.1080/10586458.2018.1434704}.

\bibitem[Liu (2004)]{liu2004comparative} LIU, Ying. A Comparative Study on Feature Selection Methods for Drug Discovery. \textit{Journal of Chemical Information and Computer Sciences}, v. 44, n. 5, p. 1823–1828, 2004. DOI: 10.1021/ci049875d.

\bibitem[Shewchuk (2024)]{Shewchuk2024} SHEWCHUK, Jonathan. \textit{Concise Machine Learning}. UC Berkeley, Class CS 189/289A Lecture Notes, 2024.

\bibitem[MacQueen (1967)]{MacQueen1967} MacQUEEN, James B. ``Some Methods for Classification and Analysis of Multivariate Observations''. In: \emph{Proceedings of the Fifth Berkeley Symposium on Mathematical Statistics and Probability}, v. 1, p. 281--297. Berkeley: University of California Press, 1967.

\bibitem[Pedregosa et al. (2011)]{ScikitLearn2011}
PEDREGOSA, F.; VAROQUAUX, G.; GRAMFORT, A.; MICHEL, V.; THIRION, B.; GRISEL, O.; BLONDEL, M.; PRETTENHOFER, P.; WEISS, R.; DUBOURG, V.; VANDERPLAS, J.; PASSOS, A.; COURNAPEAU, D.; BRUCHER, M.; PERROT, M.; DUCHESNAY, É. ``Scikit-learn: Machine Learning in Python''. \emph{Journal of Machine Learning Research}, 12, p. 2825--2830, 2011. \url{https://jmlr.org/papers/v12/pedregosa11a.html}. API e documentação disponíveis em \url{https://scikit-learn.org/stable/user_guide.html#user-guide}.

\bibitem[McNemar (1947)]{McNemarOrig1947}
McNEMAR, Quinn. ``Note on the sampling error of the difference between correlated proportions or percentages''. \emph{Psychometrika}, 12, n. 2, p. 153--157, 1947. \url{https://doi.org/10.1007/BF02295996}.

\bibitem[Smith e Ruxton (2020)]{McNemar1947}
SMITH, Matilda; RUXTON, Graeme. ``Effective use of the McNemar test''. \emph{Behavioral Ecology and Sociobiology}, 74, 133, 2020. \url{https://doi.org/10.1007/s00265-020-02916-y}.
%
% exemplo artigo com ate 5 autores
%\bibitem[]{key-2}AGOL, E.; STEFFEN, J.; SARI, R.; CLARKSON, W. (2005)
%``On detecting terrestrial planets with timing of giant planet transits''.
%\emph{Mon. Not. Roy. Astron. Soc.} 359, 567-579, \url{https://www.doi.org/10.1111/j.1365-2966.2005.08922.x}

% exemplo artigo com mais de 5 autorees
%\bibitem[]{key-4}BALLARD, S.; FABRYCKY, D.; FRESSIN, F.; CHARBONNEAU,
%D.; DESERT, J.-M.; et al. (2011) ``The Kepler-19 System: A Transiting
%2.2 R\ensuremath{\oplus} Planet and a Second Planet Detected via Transit
%Timing Variations''. \emph{Astrophys. J.} 743, id.200, \url{https://www.doi.org/10.1088/0004-637X/743/2/200} 

% exemplo artigo em proceedings
%\bibitem[]{key-1}EVERHART, E. (1985) ``An efficient integrator that
%uses Gauss-Radau spacings''. Em: \emph{IAU Colloquium 83, Dynamics
%of Comets: Their Origin and Evolution}, A. Carusi, G. B. Valsecchi
%(eds.). Astrophys. Space Sci. Lib. 115. Reidel, Dordrecht, p. %185,
%\url{https://www.doi.org/10.1007/978-94-009-5400-7_17}

\bibitem[Saito \& Rehmsmeier (2015)]{Saito2015} SAITO, Takaya; REHMSMEIER, Marc. ``The Precision-Recall Plot Is More Informative than the ROC Plot When Evaluating Binary Classifiers on Imbalanced Datasets''. \emph{PLOS ONE}, 10(3), e0118432, 2015. DOI: 10.1371/journal.pone.0118432.

% exemplo livro
%\bibitem[]{key-1}AARSETH, S. J. (2003) ``Gravitational N-body Simulations:
%Tools and Algorithms''. Cambridge University Press, Cambridge, MA,
%ISBN 978-0-521-12153-8
%
\end{thebibliography}

%%%%%%%%%%%%%%%%%%%
\appendix
\input{apendice_experimentos}
\input{apendice_hiperparametros}
\input{apendice_ambiente}

\end{document}